Contracts for unions and intersections are hard for Nickel

### 2021-06-06 paper.tex
Rework on sec 3

@@

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to disjunctions.
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

\begin{figure}[h]
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer @| (BinaryTree, BinaryTree)
\end{lstlisting}
\caption{A binary tree data type using unions}
\label{fig:union-adt-ex}
\end{figure}

In addition to internal program representations, unions are also useful to model
external data (\ref{fig:unions-use-cases}):
\begin{itemize}
    \item To represent nullable values.
    \item For plain enumerations.
    \item To support different representations of the same data.
\end{itemize}
All of which are common in configurations.

\begin{figure}[h]
% \label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : Str @| Null
\end{lstlisting}

% \label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http | `Ftp | `Sftp
\end{lstlisting}

% \label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Str}
\end{lstlisting}
\caption{Various use cases of unions}
\label{fig:unions-use-cases}
\end{figure}

In statically typed languages, and more specifically in functional programming
languages, sums (also called tagged unions, or discriminated unions) are usually
preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
syntactically separated by an associated discriminating label. Well-known
programming languages with first-class tagged unions include Haskell, OCaml,
Rust, Scala, PureScript, Elm, Swift and many more.

On the other hand, in dynamically typed languages, it is customary to directly
store different kinds of data in one variable without further ceremony. The
runtime system needs to attach type tags to values anyway, which can usually be
observed using functions like JavaScript's \verb+typeof+. Using such bare
unions\resolved{If you can observe with typeof, then it's pretty much a tagged
union, though the tag is implicit} has already been an idiom in dynamically
typed language for a long time. Being able to type and handle these unions is an
important aspect of the migration to a gradual type system. Indeed, a lot of
\resolved{This is the first time discussing gradual type systems, maybe it should
be introduced earlier, in particular how they relate to contracts
(I think is mentioned on the appendix)}
gradually typed languages either existed originally as dynamically typed
language or aim at backward compatibility with a dynamically typed base
language. This is why despite the presence of a static type system, gradually
typed languages tend to favor bare unions types which is the typed counterpart
of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. However, Nickel is a configuration
language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
untagged unions, there is a strong incentive to support untagged unions at the
type level, or at least, at the contract level.

@@

Intersections are less prevalent. An intersection is a value that belongs to
two types at the same time. A simple example is a number literal \nickel{1}
which could be accepted as both an \nickel{Int} and a \nickel{Float} without
having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
this example, the intersection \nickel{Int @& Float} can be used as a subtype of
both \nickel{Int} and \nickel{Float}.

\label{fig:intersection-subytping-ex}
\begin{lstlisting}[title={Value subtyping}]
1 : Int @& Float
\end{lstlisting}

Intersections of higher-order types corresponding to overloading
(Figure~\ref{fig:intersection-overloading-ex}).

\label{fig:intersection-overloading-ex}
\begin{lstlisting}[title={Overloading}]
incr: (Int -> Int) @& (Float -> Float) =
  fun x => x + 1
\end{lstlisting}

They can also be used to combine constraints (such as traits or interfaces), as
done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).

\label{fig:intersection-mixins-ex}
\begin{lstlisting}[language=Scala,title={Structural mixins}]
Comparator: {cmp: T -> T -> Bool}
Shower: {show: T -> String}

showSmaller : (dict: Comparator & Shower) -> T -> T -> String
\end{lstlisting}

@@

\label{fig:intersection-dual}
\begin{lstlisting}[title={Intersection as the dual of unions}]
f : (Num @| Str) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
\end{lstlisting}

\todo{Either here or the next section, show examples of unintuitive behaviour
of U and I. For example, I is not jos composition of contracts.
And union has to be treated as any one of the options}
% Things to settle: how to motivate them, how to compare them to ADTs. Purely
% dynamic approach, or typed approach too?

>>>>

% Copied from https://tex.stackexchange.com/questions/89574/language-option-supported-in-listings
\lstdefinelanguage{JavaScript}{
  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
  % keywordstyle=\color{blue}\bfseries,
  ndkeywords={class, export, boolean, throw, implements, import, this},
  ndkeywordstyle=\color{darkgray}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{purple}\ttfamily,
  % stringstyle=\color{red}\ttfamily,
  morestring=[b]',
  morestring=[b]"
}

@@

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}

@@

% Goals:
% Introduce unions and inter
% Why are they useful?
% How are they used
% How de they relate with one another

\subsection*{Unions}
Unions ($\cup$) are a fundamental building block of programs.
A union holds a value whose
shape is known to be in a fixed set of alterniteves, without exposing
which of those alternatives the value belongs to.
For example, a value that could be either a Number or a String can be
modelled using unions as \nickel{Num @| String}.
They correspond logically to disjunctions.

Statically typed languages, and more specifically functional programming
languages, provide a similar abstraction, sums
(also called tagged unions, or discriminated unions), which
also allow to model a value whose shape conforms to a fixed set of
alternatives, but requiring an explicit mark (or tag) that differentiates
to which of these alternatives it belongs to.
For example, consider the Haskell code presented on Figure 
\ref{fig:safediv-haskell}, it's a safe division that will
check if the dividend is 0, and if it is, return an
error message.
Since the function can return two different types, a sum
(\haskell{Either}) is used, and on each
of the if branches an explicit tag has to be used (either
\haskell{Left} or \haskell{Right}).

\begin{figure}[h]
\begin{lstlisting}[language=Haskell]
safeDiv :: Int -> Int -> Either String Int
safeDiv x y = 
  if y == 0 then 
    Left "Division by 0 is forbidden" 
  else
    Right (div x y)
\end{lstlisting}
\caption{Safe division implemented in Haskell}
\label{fig:safediv-haskell}
\end{figure}

This incurred verbosity certainly facilitates the work of static checks,
like typechecking, but dynamically typed languages permit much more flexibility,
and programmers don't usually expect to tag their values.
That is why, as mentioned on the Introduction, when designing
type systems for dynamically typed languages, unions become of particular
interest.
See the implementation of a safe division function on TypeScript
(Figure \ref{fig:safediv-ts}).
Neither of the if branches has to tag their return value, and the return
type of the function is modelled using unions, \typescript{string | number}.

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function safeDiv(x: number, y: number): string | number {
  if (y == 0) {
      return "Division by 0 is forbidden"
  } else {
      return x/y
  }
}
\end{lstlisting}
\caption{Safe division implemented in TypeScript}
\label{fig:safediv-ts}
\end{figure}

% \begin{figure}[h]
% \begin{lstlisting}[title={Tree-like data structure}]
% type BinaryTree =
% Integer @| (BinaryTree, BinaryTree)
% \end{lstlisting}
% \caption{A binary tree data type using unions}
% \label{fig:union-adt-ex}
% \end{figure}

Unions are not only useful to model the outcome of alternative
branches of computations, but they can be used to
formalize very common patterns of data types:
\begin{itemize}
    \item To represent nullable, or optional, values.
\begin{lstlisting}[language=nickel]
optionalValue : Str @| Null
\end{lstlisting}

    \item For plain enumerations of literals.
\begin{lstlisting}[language=nickel]
protocol : "Http" @| "Ftp" @| "Sftp"
\end{lstlisting}

    \item To support different representations of the same data.
\begin{lstlisting}[language=nickel]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Date}
\end{lstlisting}

\end{itemize}
All of which are common in configurations.

% \begin{figure}[h]
% % \label{fig:union-nullable-ex}
% \begin{lstlisting}[title={Nullable values}]
% optionalValue : Str @| Null
% \end{lstlisting}

% % \label{fig:union-enum-ex}
% \begin{lstlisting}[title={Plain enumeration}]
% protocol : `Http | `Ftp | `Sftp
% \end{lstlisting}

% % \label{fig:union-alt-ex}
% \begin{lstlisting}[title={Alternative representations}]
% person : {name : Str, age : Int}
%   @| {name : Str, dateOfBirth : Str}
% \end{lstlisting}
% \caption{Various use cases of unions}
% \label{fig:unions-use-cases}
% \end{figure}

% In statically typed languages, and more specifically in functional programming
% languages, sums (also called tagged unions, or discriminated unions) are usually
% preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
% syntactically separated by an associated discriminating label. Well-known
% programming languages with first-class tagged unions include Haskell, OCaml,
% Rust, Scala, PureScript, Elm, Swift and many more.

% On the other hand, in dynamically typed languages, it is customary to directly
% store different kinds of data in one variable without further ceremony. The
% runtime system needs to attach type tags to values anyway, which can usually be
% observed using functions like JavaScript's \verb+typeof+. Using such bare
% unions\resolved{If you can observe with typeof, then it's pretty much a tagged
% union, though the tag is implicit} has already been an idiom in dynamically
% typed language for a long time. Being able to type and handle these unions is an
% important aspect of the migration to a gradual type system. Indeed, a lot of
% \resolved{This is the first time discussing gradual type systems, maybe it should
% be introduced earlier, in particular how they relate to contracts
% (I think is mentioned on the appendix)}
% gradually typed languages either existed originally as dynamically typed
% language or aim at backward compatibility with a dynamically typed base
% language. This is why despite the presence of a static type system, gradually
% typed languages tend to favor bare unions types which is the typed counterpart
% of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

Nickel has no pre-existing untyped language that we have to
support, so we are freer of our choice.
However, Nickel is a configuration language, whose contract and type system
is intended to encode data schemas encountered in the wild. Since these schemas do not use
tags to differentiate different kinds of values, there is a strong incentive to support untagged unions at the
type level, or at least, at the contract level.

We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array<string>} or an \typescript{Array<number>},
what value could we fill \typescript{...} with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{string} or a
\typescript{number}.
That is, when using a union, we have to treat it as if it where any one
of the alternatives, in the case of the example, we should provide
it with a value that is both a \typescript{string} and a \typescript{number}.

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function addElem(list: Array<string> | Array<number>) {
  list.push(...)
}
\end{lstlisting}
\caption{Adding an element to a union of two arrays}
\label{fig:addElem}
\end{figure}

@@

Intersections ($\cap$) are less prevalent. An intersection is a value that belongs to
two types at the same time. An example is a number literal \nickel{1}
which could be accepted as both an \nickel{Int} and a \nickel{Float} without
having to perform conversions.
In that case, a value of type \nickel{Int @& Float} can be used as a value of
both type \nickel{Int} and \nickel{Float}.

% \label{fig:intersection-subytping-ex}
% \begin{lstlisting}[title={Value subtyping}]
% 1 : Int @& Float
% \end{lstlisting}

The real usability of intersections shows up when intersecting
functions, since it can be used to encode overloading.
For instance, take a look at Figure
\ref{fig:intersection-overloading-ex}, where the function
\nickel{duplicate} works both as a function to duplicate
arrays, as well as a function to duplicate strings.
This is particular useful when using unions, since it's a
good way to express that a function can deal with different
shapes of data.
For instance, on the same figure, you can use \nickel{duplicate}
over some value with type \nickel{[String] @| String}, and know
that it will work correctly.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let duplicate | ([String] -> [String]) @& (String -> String) =
  fun x => x ++ x in
let text | [String] @| String = ... in
duplicate text
\end{lstlisting}
\caption{Duplicating an array of Strings or a String}
\label{fig:intersection-overloading-ex}
\end{figure}

% \label{fig:intersection-overloading-ex}
% \begin{lstlisting}[title={Overloading}]
% incr: (Int -> Int) @& (Float -> Float) =
%   fun x => x + 1
% \end{lstlisting}

They can also be used to combine records \unsure{Better
naming}. For instance, in Figure \ref{fig:intersection-record}
two types are defined \nickel{Animal} and \nickel{Pet}, and
a variable that is compatible with both types is declared, with
type \nickel{Animal @& Pet}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal = { species: Str, race: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal & Pet = 
  { species = "Canis Lupus",
    race = "Australian Cattle Dog",
    owner = "Anonymous Author",
    name = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

% They can also be used to combine constraints (such as traits or interfaces),as
% done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).

% \label{fig:intersection-mixins-ex}
% \begin{lstlisting}[language=Scala,title={Structural mixins}]
% Comparator: {cmp: T -> T -> Bool}
% Shower: {show: T -> String}

% showSmaller : (dict: Comparator & Shower) -> T -> T -> String
% \end{lstlisting}

@@

\begin{figure}[h]
\begin{lstlisting}[language = nickel]
f : (Num @| Str) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
\end{lstlisting}
\caption{Intersection as the dual of unions}
\label{fig:intersection-dual}
\end{figure}

\resolved{Either here or the next section, show examples of unintuitive behaviour
of U and I. For example, I is not jos composition of contracts.
And union has to be treated as any one of the options.
Kindo of resolved for unions}
% Things to settle: how to motivate them, how to compare them to ADTs. Purely
% dynamic approach, or typed approach too?
